---
title: "Enrich BFS data with I14Y codelists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enrich BFS data with I14Y codelists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = curl::has_internet() &&
    requireNamespace("BFS", quietly = TRUE) &&
    requireNamespace("dplyr", quietly = TRUE) &&
    requireNamespace("stringr", quietly = TRUE) &&
    requireNamespace("readr", quietly = TRUE)
)
```

Official Swiss datasets distributed by the [Bundesamt für Statistik
(BFS)](https://www.bfs.admin.ch) often label rows in German only, using
codes from a published nomenclature (NOGA, NOMEM, NUTS, ...). The
companion [BFS R package](https://felixluginbuhl.com/BFS/) gives you the
data; **I14Y** gives you the official multilingual labels for those codes.

This vignette shows the typical pattern: **find the right codelist, download
it, join it onto a BFS dataset**.

```{r setup}
library(I14Y)
library(BFS)
library(dplyr)
library(stringr)
library(readr)
```

## 1. Find the right codelist

In the new I14Y data model, each level of a former multi-level nomenclature
is published as a **separate codelist concept**. For NOGA, that means
four concepts: `NOGA_SECTION`, `NOGA_DIVISION`, `NOGA_GROUP`, `NOGA_CLASS`.

Search for them:

```{r}
i14y_search_concept(query = "NOGA", language = "en", pageSize = 10)
```

Pick the one matching the granularity of your data. For "Wirtschaftsabteilung"
(economic division) we want `DV_NOGA_DIVISION`.

## 2. Download the codelist

```{r}
noga_division <- i14y_get_codelist(
  id = "08d94604-e058-62a2-aa25-53f84b974201" # DV_NOGA_DIVISION
) |>
  mutate(Code = as.numeric(Code))

noga_division
```

The tibble carries one column per official language (`Name_de`, `Name_fr`,
`Name_it`, `Name_en`) ready for joining.

## 3. Join to a BFS dataset

Get the monthly income by gender and economic division for 2022:

```{r, message=FALSE, warning=FALSE}
income_meta <- BFS::bfs_get_metadata(number_bfs = "px-x-0304010000_201")

income <- BFS::bfs_get_data(
  number_bfs = "px-x-0304010000_201",
  language = "de",
  query = list(
    Jahr = "2022",
    Geschlecht = c("1", "2"),
    Wirtschaftsabteilung = income_meta$values[[3]],
    "Zentralwert und andere Perzentile" = "1"
  )
)
```

Extract the NOGA code from the German label and left-join the multilingual
codelist:

```{r}
income |>
  filter(!str_detect(Wirtschaftsabteilung, "Sektor")) |>
  mutate(Code = readr::parse_number(Wirtschaftsabteilung)) |>
  left_join(noga_division, by = "Code") |>
  select(Wirtschaftsabteilung, Name_en, Name_fr, Name_it) |>
  arrange(Name_en) |>
  head(10)
```

## 4. Narrow-search inside a codelist

If you only need a few entries, search **within** a codelist concept
instead of downloading it whole. The faceted endpoint accepts a free-text
query in any of the five languages:

```{r}
i14y_search_codelist(
  id = "08d94604-e058-62a2-aa25-53f84b974201", # NOGA_DIVISION
  query = "agriculture",
  language = "fr"
)
```

And if you'd rather get the search result as a downloadable CSV:

```{r}
i14y_export_codelist_search(
  id = "08d94604-e058-62a2-aa25-53f84b974201",
  query = "agriculture",
  language = "fr",
  format = "csv"
)
```

## Concept metadata

The full concept descriptor — including version, publisher, registration
status, value type, and conformance statements — is available via
`i14y_get_concept()`:

```{r}
concept <- i14y_get_concept(
  id = "08d94604-e058-62a2-aa25-53f84b974201",
  includeCodeListEntries = FALSE
)

concept$publisher
concept$registrationStatus
concept$description$en
```

Pass `includeCodeListEntries = TRUE` if you want the entries embedded in
the same response (rarely needed when you already have
`i14y_get_codelist()`).

If you need the full canonical JSON export of a concept (useful when
archiving or feeding the concept into another system), use
`i14y_get_concept_export()`:

```{r}
concept_export <- i14y_get_concept_export(
  id = "08d94604-e058-62a2-aa25-53f84b974201"
)
str(concept_export, max.level = 1)
```
