---
title: "Discovering the I14Y catalog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovering the I14Y catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = curl::has_internet()
)
```

I14Y exposes two complementary ways to discover resources:

- `i14y_search()` — full-text search with **facets** (publisher, theme,
  registration status, ...), returns a flat tibble of resource summaries.
- `i14y_list_*()` — flat **list endpoints**, one per resource type, with
  identifier-style filters (`publisherIdentifier`, `version`, ...).

This vignette walks through both, then shows how to export catalog
metadata in DCAT for downstream tools.

```{r setup}
library(I14Y)
library(dplyr)
```

## Faceted search across all resource types

`i14y_search()` covers everything: datasets, data services, public
services, concepts, mapping tables.

```{r}
i14y_search(query = "education", language = "en") |>
  glimpse()
```

Filter by resource type:

```{r}
i14y_search(
  language = "en",
  types = c("DataService")
) |>
  glimpse()
```

Filter by publisher and theme — the values come from the I14Y vocabularies
(visible in any search result's `publisher.identifier` / `themes` fields).
For example, the Federal Statistical Office has publisher identifier `CH1`,
and the theme code `115` is "Economy":

```{r}
i14y_search(
  language = "en",
  publishers = "CH1",
  themes = "115",
  pageSize = 10
) |>
  glimpse()
```

Restrict to datasets that actually have a published structure:

```{r}
i14y_search(
  types = "Dataset",
  structure = "WithStructure",
  pageSize = 10
) |>
  glimpse()
```

## List endpoints, one per resource type

When you know the resource type and want to filter by identifier metadata
instead of full-text, use the list endpoints.

### Datasets

```{r}
i14y_list_datasets(page = 1, pageSize = 10) |>
  glimpse()
```

### Data services

```{r}
i14y_list_dataservices(page = 1, pageSize = 10) |>
  glimpse()
```

### Public services

```{r}
i14y_list_publicservices(page = 1, pageSize = 10) |>
  glimpse()
```

### Mapping tables

```{r}
i14y_list_mappingtables(page = 1, pageSize = 10) |>
  glimpse()
```

### Concepts

```{r}
i14y_list_concepts(page = 1, pageSize = 10) |>
  glimpse()
```

All of them accept the same facet-style filters: `publisherIdentifier`,
`publicationLevel` (`"Internal"` or `"Public"`), and `registrationStatus`
(one of `Incomplete`, `Candidate`, `Recorded`, `Qualified`, `Standard`,
`PreferredStandard`, `Superseded`, `Retired`). Dataset-shaped resources
additionally accept `accessRights`, and mapping tables accept
`codeSystemUri` and `version`.

## Getting a single resource

Each list endpoint has a matching get function:

```{r}
dataset <- i14y_get_dataset(id = "b902add5-9538-47ed-b663-f9fbfac92381")
dataset$title$en
dataset$registrationStatus
```

Each resource type has its own get function:

```{r}
ds_id <- i14y_list_dataservices()$id[1]
dataservice <- i14y_get_dataservice(id = ds_id)
dataservice$title$en

ps_id <- i14y_list_publicservices()$id[1]
publicservice <- i14y_get_publicservice(id = ps_id)
publicservice$title$en
```

Mapping tables expose their full metadata (source/target codelists,
publisher, version, ...) via `i14y_get_mappingtable()`:

```{r}
mt_id <- i14y_list_mappingtables()$id[1]
mt <- i14y_get_mappingtable(id = mt_id)
mt$name$en
mt$source$uri
mt$target$uri
```

The actual code-to-code relations of a mapping table (source/target/relation-type
triples) are exported with `i14y_get_mappingtable_relations()`. The default
`format = "Json"` returns a tidy data.frame; `format = "Csv"` returns the raw
CSV as a character string.

```{r}
relations <- i14y_get_mappingtable_relations(id = mt_id)
head(relations)
```

## Catalogs and DCAT export

Resources on I14Y belong to one or more **catalogs** (typically one per
federal office or cantonal organization). List them with:

```{r}
catalogs <- i14y_list_catalogs()
head(catalogs[, c("id", "title.en")], 5)
```

For a single catalog, you can fetch its metadata or download the whole
catalog as a DCAT/RDF or DCAT/TTL document:

```{r}
catalog_id <- catalogs$id[1]

catalog <- i14y_get_catalog(id = catalog_id)
catalog$title$en

dcat_ttl <- i14y_get_catalog_dcat(id = catalog_id, format = "TTL")
cat(substr(dcat_ttl, 1, 500))
```

The TTL/RDF export is the canonical format for harvesting I14Y into other
metadata systems (CKAN, opendata.swiss, custom triple-stores).

Each catalog also exposes the individual **records** (one per resource
listed in the catalog) via `i14y_get_catalog_records()`. Not every catalog
has records — many federal catalogs are currently empty — so pick one that
does, for example the NL Opendata.swiss catalog:

```{r}
records_catalog_id <- catalogs$id[catalogs$title.en == "NL Opendata.swiss"][1]
records <- i14y_get_catalog_records(id = records_catalog_id)
str(records, max.level = 1)
```

Each record points at an underlying resource (dataset, data service, ...)
through `primaryTopic.resourceId` and `primaryTopic.resourceType`, which you
can then fetch with the matching `i14y_get_*()` function:

```{r}
records[, c("id", "primaryTopic.resourceId", "primaryTopic.resourceType")] |>
  head(3)
```

The single-record endpoint `i14y_get_catalog_record(id, recordId)` exists
but currently requires authentication on the public API (HTTP 401 for
anonymous callers), so we don't show a live call here.

## Putting it together: finding all preferred-standard NOGA-related concepts

```{r}
i14y_search(
  query = "NOGA",
  language = "en",
  types = "Concept",
  conceptValueTypes = "CodeList",
  registrationStatuses = "PreferredStandard",
  pageSize = 10
)
```

From here, the steps of the
[`vignette("codelists-with-bfs")`](codelists-with-bfs.html) workflow apply:
take any `id`, call `i14y_get_codelist()`, and join.
