---
title: "Getting started with I14Y"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with I14Y}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = curl::has_internet()
)
```

The **I14Y** R package wraps the public API of the
[I14Y interoperability platform](https://www.i14y.admin.ch) of Switzerland.
The platform publishes metadata about five kinds of resources:

| Resource type    | Description                                        | Functions                                                                 |
|------------------|----------------------------------------------------|---------------------------------------------------------------------------|
| **Concept**      | Definitions and codelists                          | `i14y_search_concept()`, `i14y_get_concept()`, `i14y_get_codelist()`      |
| **Dataset**      | Tabular or non-tabular datasets                    | `i14y_list_datasets()`, `i14y_get_dataset()`, `i14y_get_dataset_structure()` |
| **DataService**  | APIs and data services                             | `i14y_list_dataservices()`, `i14y_get_dataservice()`                      |
| **PublicService**| Administrative services for citizens and business  | `i14y_list_publicservices()`, `i14y_get_publicservice()`                  |
| **MappingTable** | Mappings between codelists                         | `i14y_list_mappingtables()`, `i14y_get_mappingtable()`                    |

A single `i14y_search()` function searches across all five at once.

```{r setup}
library(I14Y)
```

## A typical workflow

Most workflows follow the same three steps:

1. **Search** for a resource, either with the faceted `i14y_search()` or with
   one of the convenience wrappers (`i14y_search_catalog()`,
   `i14y_search_concept()`).
2. **Inspect** the metadata of a single item with `i14y_get_*()`.
3. **Download** the actual payload (codelist entries, dataset structure,
   DCAT export, ...) with the matching `i14y_get_*()` / `i14y_export_*()`
   function.

### 1. Search

Search the whole catalog in English for the keyword "noga":

```{r}
i14y_search(query = "noga", language = "en")
```

The `types` column shows which resource type each row is. You can restrict
to one type:

```{r}
i14y_search(
  query = "noga",
  language = "en",
  types = "Concept"
)
```

The faceted parameters mirror the OpenAPI spec —
see `?i14y_search` for the full list (publication levels, registration
statuses, life events, business events, themes, publishers, ...).

### 2. Inspect

Pick an `id` from the search result and fetch the full metadata:

```{r}
concept <- i14y_get_concept(id = "08d94604-e058-62a2-aa25-53f84b974201")
concept$name
concept$description$en
```

Multilingual fields always carry the four official Swiss languages plus
Romansh (`de`, `fr`, `it`, `en`, `rm` when available).

### 3. Download

For a code-list concept, download the entries as a tibble:

```{r}
i14y_get_codelist(id = "08d94604-e058-62a2-aa25-53f84b974201")
```

For a dataset, the SHACL/RDF structure is available in three formats
(`JsonLd`, `Ttl`, `Rdf`):

```{r}
str(
  i14y_get_dataset_structure(
    id = "b902add5-9538-47ed-b663-f9fbfac92381",
    format = "JsonLd"
  ),
  max.level = 1
)
```

## Going further

Two further vignettes drill down into common workflows:

- `vignette("codelists-with-bfs")` — translate official BFS dataset codes
  to multilingual labels.
- `vignette("discovering-the-catalog")` — list and filter resources by type,
  publisher, theme, ...; export catalog metadata as DCAT.
